Search for Spatially Extended Fermi-LAT Sources Using Two Years of Flight Data

### 2011-06-16 extended_catalog/paper/extended_catalog.tex
Adding bibliography

@@

% some units
\newcommand{\mev}{\text{MeV}\xspace}
\newcommand{\gev}{\text{GeV}\xspace}
\newcommand{\sr}{\text{sr}\xspace}
\newcommand{\s}{\text{s}\xspace}
\newcommand{\ph}{\text{ph}\xspace}
\newcommand{\cm}{\text{cm}\xspace}
\newcommand{\tsext}{\ensuremath{\text{TS}_\text{ext}}\xspace}
\newcommand{\tsinc}{\ensuremath{\text{TS}_\text{inc}}\xspace}
\newcommand{\ts}{\text{TS}\xspace}

@@

% the program names
\newcommand{\pointlike}{{\em pointlike}\xspace}
\newcommand{\gtlike}{{\em gtlike}\xspace}
\newcommand{\gtobssim}{{\em gtobssim}\xspace}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the
Fermi Gamma-Ray Space Telescope that has been surveying the gamma ray
sky since June 2008.  Fermi is particuarly well suited for surveying
the entire Gamma-ray sky since it has a broad energy coverage ($20\mev$
and $300\gev$), a wide field of view ($\sim 2.4 \sr$) and large effective area
($\sim 8000 \cm^2$ at $>1\gev$).

The LAT has been essential in understanding a variety of astrophysical
sources including Active Galactic Nuclei, Pulsars, and Supernova Remnants.
Many of the astrophysical sources seen by the Fermi LAT, espeically
galactic Supernova Remnants and Pulsar Wind Nebulae, can be resolved at
other wavelengths and could potentially be resolved by the LAT.

studying supernova remnants, 
pulsar wind nebulae, 
galaxies,  galaxy clusters, and potentially dark matter and new physics.

Resolving spatially extended $\gev$ sources is important for studying a
variety of science topics. The most important use of spatial morphology
is in source identification. Many of the sources found in $\gev$ sources

Nevertheless, morphological studies of sources in the $\gev$ energy range using
the Fermi LAT is challenging primarily because of the significantly energy
dependent PSF and because of systematic errors in the
galactic diffuse emission. The Fermi PSF varies from over $XXXXXXXXXXXXXXX\deg$
at $100\mev$ to $XXXXXXXXXXXXXXXX\deg$ at $100\gev$ and so the higher energy photons are

The Large Area Telescope (LAT) on the Fermi Gamma-Ray Space Telescope
has been surveying the Gamma-ray sky since August 2008. The LAT has been
been essential in studying a variety of astrophysical sources including
Active Galactic Nuclie, Pulsars, Supernova Remnants. 
One of the important data productes

Using 11 months of data, the LAT produced the best-resolved survey of the sky
in the $100\mev$ to $100\gev$ energy range\cite{2FGL Paper}. The

@@

Performing morphological studies of sources in the $\gev$ energy range using
the Fermi LAT is challenging primarily because of the significantly energy
dependent PSF and because of systematic errors in the
galactic diffuse emission. The Fermi PSF varies from over $XXXXXXXXXXXXXXXX\deg$
at $100\mev$ to $XXXXXXXXXXXXXXXX\deg$ at $100\gev$ and so the higher energy photons are
significantly more important for studying the shape of sources. One could
ideally look just at the highest energy photons, but the trade off is that
for typical sources many more source events are expected at low energy.
For this reason, it is especially important to use as much of the data
as possible.

@@

\subsection{\pointlike}

@@

Instead, an alternate approach was developed for performing
a morphological analysis. An alternate maximum likelihood fitting
package called \pointlike has been developed by the Fermi
collaboration and we have expanded its functionality to fully support
analyzing spatially extended sources. \pointlike is most thoroughly
described in Matthew Kerr's Ph.D Thesis\cite{Matthew_kerr_phd_thesis}
but the important aspects will be summarized below. What differentiates
\pointlike most significantly from \gtlike is its emphasis
on speed by introducing approximations into the calculation of the
likelihood function while at the same time trying to reduce any
numerical errors that may be introduced.  Like binned \gtlike,
\pointlike bins the sky both in position and energy.  Unlike 
\gtlike, \pointlike relies on the {\em healpix} representation of spatial
bins\cite{healpix_paper} and scales the size of each spatial bin to be
smaller than the PSF. This is convenient because it no
significant information is lost due to binning at all energies and it is
always a good approximation that the model predicted counts are uniform
across a spatial bin. 
\pointlike additionaly bins the Fermi Gamma-rays by their conversion type,
which corresponds to whether the photon converted in the front or back
part of the detector. Since the PSF is smaller for front entering events,
one would expect binning each conversion type independently to improve
our sensitivity.
Furthermore, \pointlike uses a sparse matrix
representation of the spatial bins where only the bins with counts in them
are looped over. The only downside to this is that the integral of the
model predicted counts must be independently calculated.  Effectively,
\pointlike efficiently interpolates from a totally binned analysis
at low energy (where the resolution is poor and there are many counts)
to an unbinned analysis at high energy (where the resolution is very
good but there are few counts). This approach provides a significant
savings in time while introducing only small numerical error.

% XXXXXXXXXXXX describe how \pointlike calculates extension errors and how localization errors are calculated.

% XXXXXXXXXXXX note about how \pointlike is written in python = more flexible

\subsection{Extension Fitting}

We have modified \pointlike to support fitting spatially extended
sources.  To do this, we introduced a new class of diffuse sources where
the spatial and spectral information completely decouple and where
the spatial shape is parameters by simple objects (Disks, Gaussian).
The code consistently convolves the extended source shape with the point
spread function (as a function of energy) and integrates the spectrum over energy
to computed model predicted counts.

Finally, \pointlike allows the spatial parameters of an extended
source to be simulataneoulsy fit. \pointlike uses the Minuit fitting
library to perform the maximum likelihood maximization of the position
and extension of an extended source\cite{Minuit reference}.  At each
step in the fit of the shape, the spectral parameters of the entire sky
model are reoptimized. To avoid projection effects, what is directly
fit is not the longitude and latitude of the source but instead
the displacement in a rotated reference frame. The covariance matrix is
then computed and inverted to estimate the statistical errors on the
position and extension.

@@

To calculate the statistical significance of the extension of a source,
one can define a parameter $\tsext$ as twice the increase
in log likelihood between fitting a source with a
spatially extended hypothesis and a point hypothesis. 
So 
\begin{equation}
  \tsext=2\times(LL_\text{ext}-LL_\text{ps}).
\end{equation}

In going from fitting a point source to fitting a simple extended source such as
a disk, the source's width is the only degree of freedom added to the model. According to
Wilk's Theorem, the distribution of $\tsext$ for no signal (when there is a true
point source) should follow a $\chi^2_1$ distribution with one degree of freedom\cite{Wilks_Theorem}.
Wilk's Theorem only holds when the likelihood is maximized simultaneoulsy
fitting all of the parameters of model. It is for this reason why 
\pointlike
goes through such great effort to simultaneously fit the position, extension,
and spectral paramters of a source.

On the other hand, since the null hypothesis rests on the edge of
parameter space, when the extension is fixed to 0, Wilk's Theorem does
not formally hold\cite{Warnings about Wilk's Theorem Paper}.  One might
expect the actual distribution to be a half $\chi^2_1$ distribution with
one degree of freedom plus half of a delta function at 0:
\begin{equation}
  P(\ts)=\tfrac{1}{2}(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
The false detection probability can be estimated by integrating this function 
from an observed test statistic value to infinity. It is for this
particular distribution that the often quoted result holds that
$\sqrt{\ts}$ is a measure of the number of $\sigma$ of the detection.

It seems plausable that a similar situation could happen for when testing
a source for extension. Half of the time, statistical flucutations might
make the observed counts narrower thant the point spread function and
the likelhood would not improve by fitting the source with an extended
hypothesis. We tested this hypothesis below using a dedicated monte
carlo study.

One issue that we found when testing this analysis method is that
despite our best efforts, there was always a small discrepancy between
the likelihood value when fitting the source as a point source and when
fitting it as an extended source with a small extension.  This is to be
expected due to numerical innacuracies introduced by the convolution.
For most cases, this error is insignificant. It would not significanlty
bias any of the fit values or errors. But it would bias the distribution
of $\tsext$ in the null hypothesis. To correct for this,
instead of computing the likelihood in the null hypothesis by fitting
a point source, a new source model was introduced which is an extended
source with its extension fixed to a very small value. $LL_\text{ps}$
is then calculated by refitting this source using the same extended
source fitting code. It was only after introducing this change that we
were able to get broad agreement between simulations and Wilk's theorem.

% XXXXXXXXXXXXXXXXXXX NOTE ABOUT gtlike crosscheck

\subsection{Dual Localization}

@@

For a particular value of $S_x$, $S_y$, $D_x$, and $D_y$, the Poisson
likelihood is computed by converting the sum and difference coordaintes
into the rotated coordiantes, unrotating the coordiantes, moving the
two sources in the \pointlike to the new positions, and performing a new
spectral analysis at the modified position.  The function then maximizes
The Poission likelihood $L(S_x,S_y,D_x,D_y)$ as a function of the sum
and difference of the rotated coordaintes using the Minuit to fit the
sum and differenc coordiantes\cite{Minuit reference}.

After the minimization converges, we can define the 
parameter $\tsext$ as twice the 
increase
in log likelihood between fitting the region with
two point sources and fitting the region with one point source
\begin{equation}
  \tsext=2\times(LL_\text{ext}-LL_\text{ps}).
\end{equation}
Generally, \tsinc can also

% XXXXXXXXXXXXXXX \tsinc vs \tsext NOT A CALIBRATED TEST
% but still suggestive in the cases \tsinc \ll \tsext and
% \tsinc \gg \tsext.

@@

After
% XXXXXXXXXXXXXXXXXXX NOTE ABOUT gtlike crosscheck

@@

Description of comparing r68, sigma for disk, gaussian, etc.
why the plots don't look correct.

\section{Monte Carlo Validation}\label{monte_carlo_validation}

% XXXXXXXXXXXXXXXX better description of how TSext is always > 0
% but wilk's thorem says that it won't be must better if the source
% is a point source

@@

The monte carlo simulation was perforormed over a one year time interval using
a default rocking profile and an assumed livetime fraction of 0.8.
The reconstruction was perfromed with events from an energy range of
$1\gev$ to $100\gev$ using 4 energy bins per decade and using the Pass 7 Version 6
source instrument response function (IRFs) to be consistent with the following analysis.

@@

The results of the monte carlo simulation are shown in
figure~\ref{ts_ext_mc}.  The cumulative density function (CDF) is plotted
against $\tsext$ For reference, the CDF for $\chi^2_1/2$,
suggested by Wilk's Theorem, is also plotted for comparison. Overall,
there is good agreement between Wilk's theorem, although the agreement is
not perfect.  It should be noted that the discrepancy seems to be worst
for bright sources which seems to imply that numerical errors in the
way \pointlike models the PSF and performs the convolution become more
apparent for sufficiently bright sources.  Other possible reasons for
departure from Wilk's theorem might include \pointlike ignoring energy
dispersion. Energy dispersion will change the shape as a function of
energy. \pointlike may also run into possibly problems with the fitting
algorithm getting stuck in false minimum. It is important to emphasize
that if
our emperical curve lies to the left of the curve we assume to estimate
significance, we would underestimate the statistical significance of
our result. Since this is the case for most of our simluations, we are
reasonable confident that $\sqrt{\tsext}$ can be safely
used as a measure of the statistical significance of a source's extension.

\section{Extended Source Sensitivity}

@@

To calculate the LAT's sensitivity to spatially extended sources,
we simulated sources with a uniform radially symmetric
disk morphology on top of a powerlaw
Sreekumar-like isotropic spectrum.\cite{Sreekumar et al. ApJ 494 pag
523 1998}.  For a given extension and spectral index, we calculated the
sensitivity by varying the flux of a source until we found 
when $\tsext=25$. In particular, we hand picked a flux range
where $\tsext<25$ and $\tsext>25$ and simulated 10 points with
a log uniform distribution in this range and performed an extension
test for each of these fluxes. Since the measured signifciance is
a statistical which will vary from realization to realization,
we then repeated this procedure $>15$ times for each spectral
and spatial model.
By fitting a line to the flux vs $\tsext$ data, we were able to
estimage the flux at which $\tsext=25$.

@@

The LAT two year catalog presented a search for all
statistically significant sources that are detected in the $\gev$
sky\cite{2FGL_catalog}.  The catalog group's source finding method assumes
that the emission originated as point sources and what was presented was
a list of all sources fit assuming a point source spatial hyopthesis. All
published extended sources were modeled by the catalog group as extended
sources, but no attempt was made to test if other sources in the catalog
could be spatially resolved.

We performed a search of all Fermi sources found in the second Fermi LAT
catalog for extension.  We performed the extension fit using \pointlike.
To perform this search, we built a model of the sky consistent with
the two year catalog's model.   We used the same two year data set and
the same pass 7 instrument response function We used the same galactic
diffuse, isotropic diffuse, and earth limb emission model. The spectrum
of the galactic and isotropic diffuse models were refit during the
analysis. The earth limb emission was fixed to predicted value.

>>>>

% TODO:
%   Add contours to images.

@@

% some units
\newcommand{\mev}{\text{MeV}\xspace}
\newcommand{\gev}{\text{GeV}\xspace}
\newcommand{\tev}{\text{TeV}\xspace}
\newcommand{\sr}{\text{sr}\xspace}
\newcommand{\s}{\text{s}\xspace}
\newcommand{\ph}{\text{ph}\xspace}
\newcommand{\cm}{\text{cm}\xspace}
\newcommand{\tsext}{{\ensuremath{\text{TS}_\text{ext}}}\xspace}
\newcommand{\tsinc}{\ensuremath{\text{TS}_\text{inc}}\xspace}

\newcommand{\rsixeight}{{\ensuremath{\text{r}_{68}}}\xspace}

\newcommand{\tsextpointlike}{\ensuremath{\tsext_{,\pointlike}}}
\newcommand{\tsextgtlike}{\ensuremath{\tsext_{,\gtlike}}}
\newcommand{\tsextalt}{\ensuremath{\tsext_{,\alt}}}

\newcommand{\ts}{\text{TS}\xspace}
\newcommand{\glon}{\text{GLON}\xspace}
\newcommand{\glat}{\text{GLAT}\xspace}
\newcommand{\alt}{\text{alt}\xspace}

@@

% the program names
\newcommand{\pointlike}{\text{\em pointlike}\xspace}
\newcommand{\python}{\text{\em python}\xspace}
\newcommand{\gtlike}{\text{\em gtlike}\xspace}
\newcommand{\gtobssim}{\text{\em gtobssim}\xspace}
\newcommand{\minuit}{\text{\em Minuit}\xspace}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the
Fermi Gamma-Ray Space Telescope. It has been surveying the Gamma-ray
sky since June 2008.  Fermi is particularly well suited as a survey
instrument because of its to survey the broad energy coverage ($20\mev$
to $>300\gev$), a wide field of view ($\sim 2.4 \sr$) and large effective
area ($\sim 8000 \cm^2$ at $>1\gev$).

Using one year of all sky monitoring, the LAT published a catalog of
source that were significantly resolved.  Many of these sources could
be be associated with a variety of source classes including active
galactic nuclei, pulsars, galaxies, and supernova remenants.  Many of
these source classes can be spatially resolved when observed at other
frequencies and could possibly be spatially resolvable by the LAT.

In particular, the LAT has discovered a variety of spatially extended
supernova remnants including IC443 and W51C (\cite{IC443,W51C}) and pulsar
wind nebulae (\cite{MSH15-52,VelaX}). And there are other source classes
which could possibly be spatially resolved by the LAT including galaxy
clusters, dark matter, and possibly new physics.

Furthermore, a variety of spatially extended but otherwise unassociated
sources have been detected in the galacitc plane at \tev energies using
Air Cherenkov detectors. Most prominent was a survey of the galaictc
plane using H.E.S.S which discovered XXXXXXXXX spatially extended
sources\cite{HESS_plane_survey}.

Being able to spatially resolve a fermi source is important for several
reasons. Most importantly, the spatial morphology can be used to help
identify a Fermi source. Because of the large point spread function (PSF),
for any source there are often several counterparts which it could be
associated with. Finding a matching morphology can uniquly identify the
source emitting the Gamma-rays.

Because of the largely varying PSF of the LAT, the
spatial and spectral model of a source do not nicely decouple and
spatially resolving a Fermi source is also important for obtaining an
unbiased spectral model of the source.  Furthermore, using this extension
information is important for improved the overall model of the sky which
is important for obtaining a better fit of nearby sources and removing
spurious point sources which we might otherwise think we detect.

Nevertheless, morphological studies of sources in the $\gev$ energy range using
the Fermi LAT is challenging because of the significantly energy
dependent PSF and because of systematic errors in the
galactic diffuse emission. The Fermi PSF varies from over $XXXXXXXXXXXXXXX\deg$
at $100\mev$ to $XXXXXXXXXXXXXXXX\deg$ at $100\gev$ so the higher energy photons 
are significantly more important for resolving a source. On the other hand, 
we tend to run out of photons at higher energies so a detailed statistical 
test must be used to maximize our sensitivity.

@@

Performing morphological studies of sources in the $\gev$ energy range using
the Fermi LAT is challenging primarily because of the significantly energy
dependent PSF and because of systematic errors in the
galactic diffuse emission. The Fermi PSF varies from over $XXXXXXXXXXXXXXXX\deg$
at $100\mev$ to $XXXXXXXXXXXXXXXX\deg$ at $100\gev$ and so the higher energy photons are
significantly more important for studying the shape of sources. One could
ideally look at just the highest energy photons, but the trade off is that
for typical sources many more source events are expected at low energy.
For this reason, it is especially important to use as much of the data
as possible.

@@

\subsection{The \pointlike Package}

@@

Instead, an alternate approach was developed for performing
a morphological analysis. An alternate maximum likelihood fitting
package called \pointlike has been developed by the Fermi
collaboration and we have expanded its functionality to fully support
analyzing spatially extended sources. \pointlike is most thoroughly
described in Matthew Kerr's Ph.D Thesis\cite{matthew_kerr_phd_thesis}
but the important aspects will be summarized below. What differentiates
\pointlike most significantly from \gtlike is its emphasis
on speed by introducing approximations into the calculation of the
likelihood function while at the same time trying to reduce any
numerical errors that may be introduced.  Like binned \gtlike,
\pointlike bins the sky both in position and energy.  Unlike 
\gtlike, \pointlike relies on the {\em healpix} representation of spatial
bins\cite{healpix_paper} and scales the size of each spatial bin to be
smaller than the PSF. This is convenient because it no
significant information is lost due to binning at all energies and it is
always a good approximation that the model predicted counts are uniform
across a spatial bin.

\pointlike additionaly bins the Fermi Gamma-rays by their conversion
type, which corresponds to whether the photon converted in the front or
back part of the detector. Since the PSF is smaller for front entering
events, one would expect binning each conversion type independently to
improve our sensitivity.

Furthermore, \pointlike uses a sparse matrix representation of the spatial
bins where only the bins with counts in them are looped over. The only
downside to this is that the integral of the model predicted counts
must be independently calculated.  Effectively, \pointlike efficiently
interpolates from a totally binned analysis at low energy (where the
resolution is poor and there are many counts) to an unbinned analysis
at high energy (where the resolution is very good but there are few
counts). This approach provides a significant savings in time while
introducing only small numerical error.

Finally, \pointlike is written mostly in the \python programming language
using numpy and scipy to vectorize most of the computations. This provides
a reasonable tradeoff between computational efficiency and quickness
of development time. And by leaving the interface to use \pointlike as a
series of \python modules, advanced functions including extension fittion
described in section~\ref{extesion_fitting} and the dual localization
procedure described in section~\ref{dual_localization_method} could
quickly be build on top of the core modules.

\subsection{Extension Fitting}
\label{extension_fitting}

We have modified \pointlike to support fitting spatially extended
sources.  To do this, we introduced a new class of diffuse sources where
the spatial and spectral information completely decouple and where
the spatial shape is parameters by simple shapes.
The code consistently convolves the extended source shape with the point
spread function (as a function of energy) and integrates the spectrum over energy
to computed model predicted counts.

Finally, \pointlike allows the spatial parameters of an extended
source to be simulataneoulsy fit. \pointlike uses the \minuit fitting
library to perform the maximum likelihood maximization of the position
and extension of an extended source (\cite{minuit_documentation}).  At each
step in the fit of the shape, the spectral parameters of the entire sky
model are reoptimized. To avoid projection effects, what is directly
fit is not the longitude and latitude of the source but instead
the displacement in a rotated reference frame.

@@

\subsection{Extension Errors}

Errors on fit parameters are typically calculated in one of two ways. The
first method requires explicilty varying one paramter until the log of
the likelihood has fallen by a value 1/2 while simulatenously optimizing
the other fit parmaeters.  \minuit provides the {\em Simplex} function to
calculate errors this way.\cite{minuit documentation and UP parameter}
The second method involves estimating the likelihood function as being
a multivariate gaussian and calcluating the curvature of the log of
the likelihood at the peak to estimate when the function will fall by
a value 1/2. \minuit provides the {\em HESSE} algorithm to calcualte
errors this way.  The {\em HESSE} are significantly faster but make the
approximation that the log of the likelihood is perfectly gaussian.

We found that although much faster, the {\em HESSE} algorithm would
somtimes grosely understimate the errors on the localization and extension
of the source.  On the other hand, we found the {\em Simplex} algorithm
was computationally very slow and would often run into convergence
problems and fail to get reasonable errors.

As a compromise, we employed a hybrid approach to calculate
the localization and extension errors in \pointlike by making the
approximation that the covariance between the extension and location of
the source was small. On the other hand, we do not necessarily expect the
likelihood profile to be Gaussian when varying the source's extension.
So we calculate the error on the extension by fixing the position of the
source and then explicitly using {\em Simplex} to vary the extension of
the source until the log of the likelihood has fallen by 1/2.

Then, the extended source's position error is calculated using the
same code employed to calculate position errors of point sources. This
code is used to calculate the localization error of all point sources
in the catalog\cite{2FGL paper}. It works by fitting an ellipse to the
likelihood surface near the maximum and using this fit ellipse to estimate
an elliptical localization errors. By reporting two localization errors
as well as a localization error angle, this method explicitly takes care
of covariance between the two position coordinates.

% Make an extension profile plot.
% show it in figures an ddescirbeit.

\subsection{Gtlike Crosscheck}

Because of its computational speed, \pointlike is especially well
optimized for analysis that requires many iterations. This include
source localization, extension fitting, source finding, building test
statistic maps, and model comparison. On the other hand, \pointlike's
efficiency has to entail some loss of precision and so it is generally
expected that \gtlike will produce a more precise result.  Furthermore,
because it is the standard likelihood analysis package, \gtlike has been
more extensively validated and is very well trusted.

For that reason, we take our best fit source model obtained by
\pointlike with the best fit positions and extensions and refit the
spectral parameters of the model using \gtlike.  \gtlike provides a
second calculation of the likelihood which can then be used to obtain
for any source a second measure of \ts and \tsext.

We find very good agreement between the two methods. For the new extended
source candidates, Table \ref{alt_diff_model_results} shows a comparison
of \ts and \tsext computed using \pointlike and \gtlike. For the rest
of this paper, unless otherwise mentioned all \ts, \tsext, flux, and
spectral index values were obtained using \gtlike with the localization
results found by \pointlike.

\subsection{Dual Localization}
\label{dual_localization_method}

@@

For a particular value of $S_x$, $S_y$, $D_x$, and $D_y$, the Poisson
likelihood is computed by converting the sum and difference coordinates
into the rotated coordiantes, unrotating the coordiantes, moving the
two sources in the \pointlike to the new positions, and performing a new
spectral analysis at the modified position.  The function then maximizes
The Poission likelihood $L(S_x,S_y,D_x,D_y)$ as a function of the sum
and difference of the rotated coordaintes using the Minuit to fit the
sum and differenc coordiantes (\cite{minuit_documentation}).

After the minimization converges, we can define the parameter $\tsext$
as twice the increase in log likelihood between fitting the region with
two point sources and fitting the region with one point source
\begin{equation}
  \tsext=2\times(LL_\text{ext}-LL_\text{ps}).
\end{equation}
Generally, \tsinc can be directly compared to \tsext to see wheter two
point sources are prefered over one extended source. Unfortunatly,
performing a likelihood ratio test between two unnested models
is not a calibrated test and so there is no straightforward way
to convert the change in likelihood into a significance~\ref{
http://arxiv.org/abs/astro-ph/0201547 }. Nevertheless, we find this
ratio, at least in the extreme case where $\tsinc \ll \tsext$ or
$\tsinc\gg\tsext$ to be suggestive. For all the spatally extended
candidates found, we computed both \tsext and \tsinc.

@@

@@

% XXXXXXXXXXXXXXX This section needs work

When running \pointlike, we have to pick a spatial model for an extended
source. Our spatial model is then convolved with the PSF to create the
probability density function (PDF) that is is fit to the obersved counts.
Possible choices for spatial models include a Two dimensional gaussian
source (defined as a Gaussian in $x$ times a Gaussian in $y$):
\begin{equation}
  \text{PDF}(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk:
\begin{equation}
  \text{PDF}(x,y)=\tfrac{1}{\pi\sigma^2}\delta\left(\sqrt{x^2+y^2}<\sigma\right).
\end{equation}

Although these shapes look significantly different from eachother, the
convolution of these spatial models of typtical sizes 
with the PSF will produce PDFs that are very similar.

This is demonstrated in figure~\ref{compare_disk_gauss}.  The plot shows
(in black) the PSF that would be observed for a source of spectral index
2. The dashed red (blue) line shows the probabability density function
for a Gaussian (uniform disk) of width $0.5\deg$.  The solid red (blue)
line shows the probability density function for the same Gaussian (uniform
disk) convolved with the PSF of an index 2 source.  
figure~\ref{compare_disk_gauss} shows the analysis using an energy range
of 1\gev to 100\gev whereas the right plot shows an analysis using an
energy range from 10\gev to 100\gev.

This plot shows that significantly different extended source shapes
look very similar after smearing with the PSF. The convolved shape of
the disk and Gaussian admittedly do look more different when restrict
ourselves to higher energy where the PSF is narrower, but we do not
excpet to have much luch learning about the exact nature of the
extended source shape. For that reason, in this paper we always use
the uniform disk spatial model to emphaizie our ignorance to the true
shape of the extended sources we are studying.

To get the convolved disk and Gaussian models looked so similar in
figure~\ref{compare_disk_gauss}, we had to pick the extension of the two
sources so that they each predicted a 68\% containment radius of 0.5\deg.

The 68\% containment radius for a spatial model is
the radius at which 68\% of the intensity is enclosed.
It can be computed by a straightforward integral and we
find that $\rsixeight_\text{,disk}=0.82\sigma$ whereas
the $\rsixeight_\text{,Gaussian}=1.51\sigma$. Therefore, in
figure~\ref{compare_disk_gauss} the uniform disk has an extension
of $\sigma=0.41\deg$ whereas the Gaussian has an extension of
$\sigma=0.75\deg$. Even though these sources have quite different
values of the extension $\sigma$, they look almost identical when convolved
with the PSF.

Finally, we note that comparing observed morphologies to 
counts maps can be misleading.  This effect can best be seen
in figure~\ref{compare_r68} where we have simulted a
extended source with 
a spatial model that is a uniform disk of extension 0.5\deg,
integral flux between 1\gev and 100\gev of $3e-8\ph/\cm^2/s$,
and a spectral index of 2.

To reduce Poisson noise, figure~\ref{compare_r68} shows a counts map of
this simulated source smoothed with a 0.1\deg Gaussain kernel.  Overlaied
on the plot is the results of fitting this simulated source with a uniform
disk and a Gaussian spatial model.  The solid white line represents the
fit extension $\sigma_\text{disk}=0.496$ for the disk which was found to
be in very good agreemen with the simulated value.  The solid black line
represenst the fit extension $\sigma$ for the Gaussian which was found to
be $\sigma_\text{Gaussian}=0.2627$.  The white dashed line is the 68\%
containment radius of the fit Disk shape and the black dashed line is
the 68\% containment radius of the fit Gaussian shape.  Even though the
fit $\sigma$ for the disk and uniform Gaussian are different, the fit
68\% continment radius are very similar ($\rsixeight_\text{,disk}=0.41$
whereas $\rsixeight_\text{,Gaussian}=0.40$).

There are three important conclusions to take from this. The first is
that depending upon what spatial model you pick and what parameter you
choose to overlay on the smoothed counts ($\sigma$ or $rsixeight$),
you will get very different looking shapes even though the all lead
to a PDF which is very similar. And since we are not very sensitive
to the particular morphology of extended sources that we are seeing,
there is no correct choice for which shape to use. For the following
analysis, we always use a uniform disk as our spatial model because
of its conceptual simplicity and to emphasize our ignorance to the
true morphology. And for subsequent smoothed counts maps, we overlay
the $\sigma$ because that shows directly the edge of the shape we fit.
But due to smearing of the PSF, it has the unfotunatly concequence that
the extended source shape always (including in the simulated image)
looks larger than in the counts map.

The second conclusion we make is that because a counts map is smeared
out by the PSF and depends strongly on the particular color bar used, it
is very difficult to determine directly from a counts map a reasonable
spatial model that fits the data.  It is much better to obtain the
spatial model by a direct fit of a spatial parameters of a selected
model to the data.  The third conclusion we make is that when comparing
fit extensions to those at other wavelenghts (espeically in \tev), it is
important to compare fits of consistent spatial models. If this is not
possible, it is impotant to at least compare indirectly the $rsixeight$
for the different spatial models.

\section{LAT Extended Source Validation}

\subsection{Significance of Extension}

To calculate the statistical significance of the extension of a source,
one can define a parameter $\tsext$ as twice the increase in log
likelihood between fitting a source with a spatially extended hypothesis
and a point hypothesis:
\begin{equation}
  \tsext=2\times(LL_\text{ext}-LL_\text{ps}).
\end{equation}
Of course, these are nested hypotehsis since a point sources is just an
extended source in the limit that the extension goes to 0. So since we
are just adding a paramter, it should always be true that $\tsext > 0$
We would expect that if we were really testing a point source that
\tsext would not be very large since a point model would fit about
as good as anytthing else. Therefore, we can consider a source to be
extended (and reject the null hyopthesis) if \tsext is large enough.
Since going from fitting a point soue to fitting a simple extended
source introduces only one degree of freedom (the source's extension)
Wilks' Theorem says that the distribution of $\tsext$ wen testin
point sources should follow a $\chi^2_1$ distribution with one degree
of freedom (\cite{wilks_theorem}).

On the other hand, since the null hypothesis rests on the edge of
parameter space, when the extension is fixed to 0, Wilks' Theorem does
not formally hold (\cite{warn_wilks_theorem}).  One might
expect the actual distribution to be a half $\chi^2_1$ distribution with
one degree of freedom plus half of a delta function at 0:
\begin{equation}
  P(\ts)=\tfrac{1}{2}(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
Chernoff proved that given certain assumtipons, this is
the ditribution that would hold when the null hypothesis
lies on the edge of parameter space for a particular parameter~\ref{chernof}.
The false detection probability can be estimated by integrating this function 
from an observed test statistic value to infinity. It is for this
particular distribution that the often quoted result holds that
$\sqrt{\ts}$ is a measure of the number of $\sigma$ of the detection.

It should be noted that Wilks' Theorem only holds when the likelihood
is maximized simultaneoulsy fitting all of the parameters of model. It
is for this reason why \pointlike goes through such great effort to
simultaneously fit the position, extension, and spectral paramters of
a source.

It seems plausable that a similar situation could happen when testing
a source for extension. Half of the time, statistical flucutations might
make the observed counts narrower thant the point spread function and
the likelhood would not improve by fitting the source with an extended
hypothesis. We tested this hypothesis below using a dedicated monte
carlo study.

One issue that we found when testing this analysis method is that
despite our best efforts, there was always a small discrepancy between
the likelihood value when fitting the source as a point source and when
fitting it as an extended source with a small extension.  This is to be
expected due to numerical innacuracies introduced by the convolution.
For most cases, this error is insignificant. It would not significanlty
bias any of the fit values or errors. But it would bias the distribution
of $\tsext$ in the null hypothesis. To correct for this,
instead of computing the likelihood in the null hypothesis by fitting
a point source, a new source model was introduced which is an extended
source with its extension fixed to a very small value. $LL_\text{ps}$
is then calculated by refitting this source using the same extended
source fitting code. It was only after introducing this change that we
were able to get broad agreement between simulations and Wilks' theorem.

\subsection{Monte Carlo Validation} \label{monte_carlo_validation}

@@

The monte carlo simulation was perforormed over a one year time interval using
a default rocking profile and an assumed a represntative
livetime fraction of 0.8.
The reconstruction was perfromed with events from an energy range of
$1\gev$ to $100\gev$ using 4 energy bins per decade and using the Pass 7 Version 6
source instrument response function (IRFs) to be consistent with the following analysis.

@@

The results of the monte carlo simulation are shown in
figure~\ref{ts_ext_mc}.  The cumulative density function (CDF) is plotted
against $\tsext$ For reference, the CDF for $\chi^2_1/2$,
suggested by Wilks' Theorem, is also plotted for comparison. Overall,
there is good agreement between Wilks' theorem, although the agreement is
not perfect.  It should be noted that the discrepancy seems to be worst
for bright sources which seems to imply that numerical errors in the
way \pointlike models the PSF and performs the convolution become more
apparent for sufficiently bright sources.  Other possible reasons for
departure from Wilks' theorem might include \pointlike ignoring energy
dispersion. Energy dispersion will change the shape as a function of
energy. \pointlike may also run into possibly problems with the fitting
algorithm getting stuck in false minimum. It is important to emphasize
that if
our emperical curve lies to the left of the curve we assume to estimate
significance, we would underestimate the statistical significance of
our result. Since this is the case for most of our simluations, we are
reasonable confident that $\sqrt{\tsext}$ can be safely
used as a measure of the statistical significance of a source's extension.

\section{Extended Source Sensitivity}\label{extension_sensitivity}

@@

To calculate the LAT's sensitivity to spatially extended sources,
we simulated sources with a uniform radially symmetric
disk morphology on top of a powerlaw Sreekumar-like isotropic
spectrum.\cite{Sreekumar et al. ApJ 494 pag 523 1998}.  The monte carlo
simulation was perforormed over a one year time interval using a default
rocking profile and an assumed a represntative livetime fraction of 0.8.
For a given extension and spectral index, we calculated the
sensitivity by varying the flux of a source until we found 
when $\tsext=25$. In particular, we hand picked a flux range
where $\tsext<25$ and $\tsext>25$ and simulated 10 points with
a log uniform distribution in this range and performed an extension
test for each of these fluxes. Since the measured signifciance is
a statistical which will vary from realization to realization,
we then repeated this procedure $>15$ times for each spectral
and spatial model.
By fitting a line to the flux vs $\tsext$ data, we were able to
estimage the flux at which $\tsext=25$.

@@

The LAT two year catalog presented a search for all
statistically significant sources that are detected in the $\gev$
sky\cite{2FGL_catalog}.  The catalog group's source finding method assumes
that the emission originated as point sources and what was presented was
a list of all sources fit assuming a point source spatial hypothesis. 
Published extended sources were modeled by the catalog group as 
being extended, but no attempt was made to test if other sources in the catalog
could be spatially resolved.

We performed a search of all Fermi sources found in the second Fermi LAT
catalog for extension.  To perform this search, we built a model of the
sky consistent with the two year catalog's model.   We used the same two
year data set running from XXXXXXXXXXX MET to XXXXXXXXXXX MET and we used
the same pass 7 instrument response function.  We used the same galactic
diffuse, isotropic diffuse, and earth limb emission model. The spectrum of
the galactic and isotropic diffuse models were refit during the analysis
while the earth limb emission was fixed to it's predicted value.

We performed the extension fit using \pointlike.  For each source, we
used a circular $10\deg$ region of interest centered on our source and
using 4 bins uniformly in each logarithm of energy.  We included all
background catalog sources within $15\deg$ of the source of interest
and refit the spectral parameters of sources within $2\deg$ of our source.
